---
title: "Analyzing the Florida Museum Historic Ceramic Type Collection"
format: html
editor: visual
---

```{r}
library(tidyverse)
library(tidyvec)
library(here)
```

## Embeddings
```{r}
# Create a siglip embedder (Python dependencies auto-provisioned on first use)
siglip_embedder <- embedder_hf("google/siglip-so400m-patch14-224",
                             modality = "multimodal",
                             device = 'mps')

# Get paths to example images included with the package
img_paths <- list.files('../../ceramic-explorer/data/data-derived/image_front_clean',
                        full.names = TRUE)

# Load metadata
metadata <- read_tsv(here('../ceramic-explorer/data/data-raw/histarch_photos.tsv'),
                     show_col_types = FALSE) %>%
  select(PhotoID, specimen_no, ceramic_category, type_name,
         glaze, surface_decoration, production_origin,
         vessel_forms, rim_forms, base_forms, motif_1, background_color) %>%
  mutate(across(everything(), tolower),
         PhotoID = as.numeric(PhotoID))

# Create an image collection with metadata
images <- tibble(
  id = parse_number(str_split_i(img_paths, "/", 7)),
  path = img_paths) %>%
  left_join(metadata, by = c("id" = "PhotoID")) %>%
  vec(embedding_fn = siglip_embedder) %>%
  embed(content_column = "path")
```

## Explore the Collection

```{r}
# Collection composition
cat("Total ceramics:", nrow(images), "\n")
cat("Unique ceramic categories:", n_distinct(images$ceramic_category), "\n")
cat("Unique production origins:", n_distinct(images$production_origin), "\n")
```

```{r}
# Distribution of ceramic categories
images %>%
  count(ceramic_category, sort = TRUE) %>%
  filter(!is.na(ceramic_category), ceramic_category != "null") %>%
  head(10) %>%
  ggplot(aes(x = reorder(ceramic_category, n), y = n)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Top 10 Ceramic Categories",
    x = NULL,
    y = "Count"
  ) +
  theme_minimal()
```

```{r}
# Distribution of production origins
images %>%
  count(production_origin, sort = TRUE) %>%
  filter(!is.na(production_origin), production_origin != "null") %>%
  head(10) %>%
  ggplot(aes(x = reorder(production_origin, n), y = n)) +
  geom_col(fill = "coral") +
  coord_flip() +
  labs(
    title = "Top 10 Production Origins",
    x = NULL,
    y = "Count"
  ) +
  theme_minimal()
```

## Visualize Embedding Space

```{r}
# Visualize our image embeddings colored by category
viz_embeddings(images,
               method = "umap",
               color = "ceramic_category") +
  labs(title = "Ceramic Embeddings in 2D Space",
       subtitle = "Colored by ceramic category")
```

```{r}
write_vec(images, 'flmnh_histarch.qs')
```

## Load Collection from Disk

When loading a saved collection, you can optionally restore the embedding function
if you need to query with new text or images.

```{r}
# Load without embedding function (for analysis with existing embeddings)
images <- read_vec('flmnh_histarch.qs')

# Or load WITH embedding function (for querying with new text/images)
images <- read_vec('flmnh_histarch.qs', embedding_fn = siglip_embedder)
```

## Semantic Search Examples

Multimodal models enable searching with natural language descriptions.

```{r}
# Search for blue and white decorated ceramics
nearest(images, "blue and white decorated pottery", n = 9) |>
  viz_images(path_column = "path",
             label_columns = c("id", "ceramic_category", "production_origin"))
```

```{r}
# Search for glazed pottery
nearest(images, "glazed ceramic pottery", n = 9) |>
  viz_images(path_column = "path",
             label_columns = c("id", "glaze", "ceramic_category"))
```

```{r}
# Search for Spanish colonial majolica
nearest(images, "spanish colonial majolica pottery", n = 9) |>
  viz_images(path_column = "path",
             label_columns = c("id", "ceramic_category", "production_origin"))
```

## Evaluate Retrieval Quality

Use tidyvec's evaluation metrics to measure how well embeddings capture ceramic similarity.

```{r}
# Calculate MAP for ceramic_category
map_score <- calculate_map(images, "ceramic_category", k = 25)
baseline <- calculate_baseline(images, "ceramic_category")

cat("Mean Average Precision (ceramic_category):", round(map_score, 3), "\n")
cat("Baseline (random retrieval):", round(baseline, 3), "\n")
cat("Improvement over baseline:", round((map_score - baseline) / baseline * 100, 1), "%\n")
```

```{r}
# Evaluate multiple variables
eval_results <- evaluate_retrieval(
  images,
  variables = c("ceramic_category", "production_origin", "glaze"),
  k = 25
)

eval_results |>
  arrange(desc(map)) |>
  mutate(
    improvement = (map - baseline) / baseline * 100
  ) |>
  select(variable, map, baseline, improvement)
```

## Tidyverse Integration

tidyvec works seamlessly with dplyr and other tidyverse tools.

```{r}
# Filter to majolica, then search within that subset
images |>
  filter(ceramic_category == "majolica") |>
  nearest("colorful floral decoration", n = 6) |>
  select(id, ceramic_category, production_origin, type_name, similarity)
```

```{r}
# Search and group results by production origin
nearest(images, "glazed pottery", n = 50) |>
  filter(!is.na(production_origin), production_origin != "null") |>
  group_by(production_origin) |>
  summarize(
    count = n(),
    mean_similarity = mean(similarity)
  ) |>
  arrange(desc(count))
```

```{r}
# Find porcelain, examine glaze types
images |>
  filter(ceramic_category == "porcelain") |>
  count(glaze, sort = TRUE) |>
  filter(!is.na(glaze), glaze != "null")
```
